---
title: "AE 04: Adjusting scales for World Bank indicators"
author: "Christina Feng (cef229)"
format: html
execute:
  warning: false
  message: false
---

```{r}
#| label: packages

library(tidyverse)
library(viridis)
library(scales)

options(scipen = 999)   # avoid printing in scientific notation
theme_set(theme_minimal()) # different default theme
```

# Data: World economic measures

The [World Bank](https://data.worldbank.org/) publishes a rich and detailed set of socioeconomic indicators spanning several decades and dozens of topics. Here we focus on a few key indicators for the year 2021.

-   `gdp_per_cap` - [GDP per capita (current USD)](https://data.worldbank.org/indicator/NY.GDP.PCAP.CD)
-   `pop` - [Total population](https://data.worldbank.org/indicator/SP.POP.TOTL)
-   `life_exp` - [Life expectancy at birth, total (years)](https://data.worldbank.org/indicator/SP.DYN.LE00.IN)
-   `female_labor_pct` - [Labor force, female (% of total labor force)](https://data.worldbank.org/indicator/SL.TLF.TOTL.FE.ZS)
-   `income_level` - [Classification of economies based on national income levels](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups)

The data is stored in `wb-indicators.rds`. To import the data, use the `read_rds()` function.

```{r}
#| label: import-data
#| message: false

world_bank <- read_rds("data/wb-indicators.rds")
```

# Part 1: Transforming axes

Is there a relationship between a country's per capita GDP and life expectancy? Let's explore this relationship using a scatterplot.

```{r}
#| label: gdp-life-scatter

ggplot(data = world_bank, mapping = aes(x = gdp_per_cap, y = life_exp)) +
  geom_point() +
  labs(
    title = "Countries with higher GDP tend to have higher life expectancy",
    x = "GDP per capita (current USD)",
    y = "Life expectancy at birth (years)"
  )
```

Seems like there is an association, but the relationship is not linear. Let's try a log transformation on the $x$-axis to see if that helps.

**Your turn:** Log-transform the $x$-axis by mutating the original column prior to graphing.

::: callout-note
By default, `log()` computes **natural logarithms** (base-$e$). To compute base-10 logarithms, use `log10()`.
:::

```{r}
#| label: log-transform-raw

# add code here

ggplot(data = world_bank, mapping = aes(x = log10(gdp_per_cap), y = life_exp)) +
  geom_point() +
  labs(
    title = "Countries with higher GDP tend to have higher life expectancy",
    x = "GDP per capita (current USD)",
    y = "Life expectancy at birth (years)"
  )
```

**Your turn:** Now log-transform the $x$-axis by using the original per capita GDP measure and an appropriate `scale_x_*()` function.

```{r}
#| label: log-transform-scale

# add code here

ggplot(data = world_bank, mapping = aes(x = gdp_per_cap, y = life_exp)) +
  scale_x_log10() +
  geom_point() +
  labs(
    title = "Countries with higher GDP tend to have higher life expectancy",
    x = "GDP per capita (current USD)",
    y = "Life expectancy at birth (years)"
  )
```

**Your turn:** Which is more interpretable, and why?

*The second plot is more interpretable because the x-axis has the GDP per capita in intervals of logarithms (1000, 10000, and 100000) , while the first plot simply has the logarithmic scale value (3, 4, 5).*

# Part 2: Customize scales

Let's consider the relationship between female labor participation and per capita GDP. We'll use the `income_level` variable to color the points and provide context on the overall wealth of the countries.[^1]

[^1]: Note that the income level is based on the GNI per capita, which is strongly correlated with GDP per capita, but not exactly the same.

## Step 1: Base plot

First, let's generate a color-coded scatterplot with a single smoothing line.

```{r}
#| label: base-plot
ggplot(data = world_bank, mapping = aes(x = female_labor_pct, y = gdp_per_cap)) +
  geom_point(mapping = aes(color = income_level)) +
  geom_smooth(se = FALSE)
```

## Step 2: Your turn

Now, let's modify the scales to make the chart more readable. Log-transform the $y$-axis and format the labels so they are explicitly identified as percentages and currency.

```{r}
#| label: format-scales

# add code here
ggplot(data = world_bank, mapping = aes(x = female_labor_pct, y = gdp_per_cap)) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_log10(labels = label_currency(scale_cut = cut_short_scale())) +
  geom_point(mapping = aes(color = income_level)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Female labor (percentage of total workforce)",
    y = "GDP per capita (current USD)"
  )
```

## Step 3: Your turn

Add human-readable labels for the title, axes, and legend.

```{r}
#| label: add-labels

# add code here
ggplot(data = world_bank, mapping = aes(x = female_labor_pct, y = gdp_per_cap)) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_log10(labels = label_currency(scale_cut = cut_short_scale())) +
  geom_point(mapping = aes(color = income_level)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Relationship between Female labor participation and GDP",
    x = "Female labor (percentage of total workforce)",
    y = "GDP per capita (current USD)",
    color = "Income Level"
  )
```

## Step 4: Your turn

Use the {viridis} color palette for `income_level`.

::: callout-tip
The bright yellow at the end of the palette is hard on the eyes. You can condense the hue at which the color map ends using the `end` argument to the appropriate `scale_color_*()` function.
:::

```{r}
#| label: viridis

# add code here
ggplot(data = world_bank, mapping = aes(x = female_labor_pct, y = gdp_per_cap)) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_log10(labels = label_currency(scale_cut = cut_short_scale())) +
  scale_color_viridis_d(end = .8) +
  geom_point(mapping = aes(color = income_level)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Relationship between Female labor participation and GDP",
    x = "Female labor (percentage of total workforce)",
    y = "GDP per capita (current USD)",
    color = "Level of Income"
  )
```

## Step 5: Your turn

Double-encode the `income_level` variable by using both color and shape to represent the same variable. Condense the guides so you use a single legend.

```{r}
#| label: double-encode

# add code here
ggplot(data = world_bank, mapping = aes(x = female_labor_pct, y = gdp_per_cap)) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_log10(labels = label_currency(scale_cut = cut_short_scale())) +
  scale_color_viridis_d(end = .8) +
  geom_point(mapping = aes(color = income_level, 
                           shape = income_level)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Relationship between Female labor participation and GDP",
    x = "Female labor (percentage of total workforce)",
    y = "GDP per capita (current USD)",
    color = "Level of income",
    shape = "Level of income"
  )
```

## Step 6: Your turn

It's annoying that the order of the values in the legend are opposite from how the income levels are ordered in the chart. Reverse the order of the values in the legend so they correspond to the ordering on the $y$-axis.

```{r}
#| label: reverse-legend

# add code here
ggplot(data = world_bank, mapping = aes(x = female_labor_pct, y = gdp_per_cap)) +
  scale_x_continuous(labels = label_percent(scale = 1)) +
  scale_y_log10(labels = label_currency(scale_cut = cut_short_scale())) +
  scale_color_viridis_d(end = .8,
                        guide = guide_legend(reverse = TRUE)) +
  geom_point(mapping = aes(color = income_level, 
                           shape = income_level)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Relationship between Female labor participation and GDP",
    x = "Female labor (percentage of total workforce)",
    y = "GDP per capita (current USD)",
    color = "Level of income",
    shape = "Level of income"
  )

```
